# Age-specific rates and probabilities {#week2}

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(DT)
library(ggplot2)
library(HMDHFDplus)
library(keyring)
library(latex2exp)
library(LexisPlotR)
library(plotly)
library(purrr)
library(scales)
library(slickR)
library(svglite)
library(tidyverse)
```

For most human populations, demographic rates vary strongly by age.

```{r load_hmd_asr, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
# Functions to download a specified dataset item for a single county
read_hmd_country <- function(CNTRY, item) {
  HMDHFDplus::readHMDweb(
    CNTRY = CNTRY,
    item = item,
    username = keyring::key_list("human-mortality-database")$username,
    password = keyring::key_get(
      service = "human-mortality-database",
      username = keyring::key_list("human-mortality-database")$username
    )
  )
}
read_hfd_country <- function(CNTRY, item) {
  HMDHFDplus::readHFDweb(
    CNTRY = CNTRY,
    item = item,
    username = keyring::key_list("human-fertility-database")$username,
    password = keyring::key_get(
      service = "human-fertility-database",
      username = keyring::key_list("human-fertility-database")$username
    )
  )
}

# Help function to list the available countries
countries_hmd <- HMDHFDplus::getHMDcountries()
countries_hfd <- HMDHFDplus::getHFDcountries()

# Download a dataset iteratively for all countries using purrr::map()
# In this case, age-specific mortality in 1-year periods x 1-year age groups
# for all 1-year periods available
mx_1x1 <- countries_hmd %>%
  # Returns a list of data.frames, adding a column for country code to each
  purrr::map(function(country) {
    read_hmd_country(country, "Mx_1x1") %>%
      dplyr::mutate(country = country)
  }) %>%
  # Combines the data.frames into a single data.frame
  dplyr::bind_rows() %>%
  dplyr::filter(Year == 2020, Age <= 90)
number_hmd_countries <- mx_1x1 %>%
  dplyr::distinct(country) %>%
  dplyr::pull(country) %>%
  length()

# Repeat for asfr
asfr <- countries_hfd %>%
  # Returns a list of data.frames, adding a column for country code to each
  purrr::map(function(country) {
    read_hfd_country(country, "asfrRR") %>%
      dplyr::mutate(country = country)
  }) %>%
  # Combines the data.frames into a single data.frame
  dplyr::bind_rows() %>%
  dplyr::filter(Year == 2019)
number_hfd_countries <- asfr %>%
  dplyr::distinct(country) %>%
  dplyr::pull(country) %>%
  length()
```

Below is the **age pattern of mortality** from ages 0 to 90 in `r number_hmd_countries` countries in 2020^[From the Human Mortality Database: https://www.mortality.org/]

```{r hmd_asmr_plot_by_country, echo=FALSE, message=FALSE, warning=FALSE}
mx_1x1 %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = Age, y = Total, group = country) +
  ggplot2::geom_line() +
  ggplot2::scale_y_continuous(
    trans = log_trans(),
    breaks = scales::trans_breaks("log", function(x) exp(x)),
    labels = scales::trans_format("log", math_format(e^.x))
  ) +
  ggplot2::xlab("1-year age group") +
  ggplot2::ylab("Period\nage-specific\nmortality rate\n(log-scale)") +
  ggplot2::theme_minimal() +
  ggplot2::theme(panel.grid = element_blank(),
                 axis.title.y = element_text(angle = 0))
```

Now look at the **age pattern of fertility** from ages 12 to 55 in `r number_hfd_countries` countries in 2019^[From the Human Fertility Database: https://www.humanfertility.org/]

```{r hmd_asfr_plot_by_country, echo=FALSE, message=FALSE, warning=FALSE}
asfr %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = Age, y = ASFR, group = country) +
  ggplot2::geom_line() +
  ggplot2::xlab("1-year age group") +
  ggplot2::ylab("Period\nage-specific\nfertility rate") +
  ggplot2::theme_minimal() +
  ggplot2::theme(panel.grid = element_blank(),
                 axis.title.y = element_text(angle = 0))
```

## Period age-specific rates

General definition of a period age-specific rate^[OMG I hate this notation!]:

$$
{}_{n}R_{x}[0,T] = \frac{
  \textsf{Number of occurrences in the age range } x \textsf{ to } x + n \textsf{ between time } 0 \textsf{ and } T
}{
  \textsf{Number of person-periods lived in the age range } x \textsf{ to } x + n \textsf{ between time } 0 \textsf{ and } T
}
$$

* Mortality pattern plot above shows **age-specific mortality rates** ${}_{n}M_{x}[0,T]$
* Birth pattern plot above shows **age-specific fertility rates** ${}_{n}F_{x}[0,T]$
* Interval length $n$ given in *exact years*
    * **Example:** ${}_{5}M_{30}[0,T]$ is mortality rate between exact ages $30.\overline{0000}$ and $34.\overline{9999}$
    
### Crude rates as a function of age-specific rates and age structure {.unnumbered}

Consider crude death rate $CDR[0,T] = D/N$ measured from death counts $D$ and mid-period population size^[Estimate of person-years] $N$:

$$
\begin{align}
CDR &= \frac{D}{N}
     = \frac{\sum_{x=0}^{\infty} {}_{n}D_{x}}{N}
     = \frac{\sum_{x=0}^{\infty} \frac{{}_{n}D_{x}}{{}_{n}N_{x}}{}_{n}N_{x}}{N} \\
    &= \sum_{x=0}^{\infty} \frac{{}_{n}D_{x}}{{}_{n}N_{x}} \cdot \frac{{}_{n}N_{x}}{N} \\
    &= \sum_{x=0}^{\infty} {}_{n}M_{x}\cdot {}_{n}C_{x} \\
    &= \textsf{Sum over age groups of }
       \begin{pmatrix}
        \textsf{Age-specific} \\
        \textsf{mortality rate}
       \end{pmatrix}
       \times \begin{pmatrix}
        \textsf{Proportion in} \\
        \textsf{age group}
       \end{pmatrix}
\end{align}
$$

Points to keep in mind about this equation:

* ${}_{n}C_{x} = {}_{n}N_{x} / N$ is proportion of population aged $x$ to $x+n$
* $C_i$ can also be interpreted as proportion of person-years lived by those aged $x$ to $x+n$
* $\sum_{x=0}^\infty {}_{n}C_{x} = 1$
* The set of $C_i$ defines the population's **age structure**
* In practice, you sum up to the oldest age group in the population
* With age intervals of inequal length, you can index age groups by $i$:

$$CDR = \sum_{i}^{\infty} M_{i}\cdot C_{i}$$

This result can resolve confusing cross-spatial and cross-temporal comparisons.

**FOR EXAMPLE**

```{r us_asmr_paradox_setup, echo=FALSE, message=FALSE, warning=FALSE}
deaths <- read_hmd_country("USA", "Deaths_1x1") %>%
  dplyr::rename(Deaths = Female) %>%
  dplyr::select(Year, Age, Deaths) %>%
  dplyr::filter(Year %in% c(1933, 2019))
exposures <- read_hmd_country("USA", "Exposures_1x1") %>%
  dplyr::rename(Exposures = Female) %>%
  dplyr::select(Year, Age, Exposures) %>%
  dplyr::filter(Year %in% c(1933, 2019))
asmr_us <- deaths %>%
  dplyr::inner_join(exposures) %>%
  dplyr::mutate(m = Deaths / Exposures) %>%
  dplyr::group_by(Year) %>%
  dplyr::mutate(p = Exposures / sum(Exposures)) %>%
  dplyr::ungroup()
cdr_us <- asmr_us %>%
  dplyr::group_by(Year) %>%
  dplyr::summarize(cdr = sum(m * p)) %>%
  dplyr::ungroup() %>%
  dplyr::select(Year, cdr)
```

Below are the crude death rates in the U.S. for women in 1933 vs. 2019^[Again calculated from the Human Mortality Database]. CDR is higher in 1933. Makes sense. But 1933 was the height of the Great Depression. I'd expect a bigger gap. WTF?

```{r us_asmr_paradox_cdrs, echo=FALSE, message=FALSE, warning=FALSE}
DT::datatable(
  dplyr::mutate(cdr_us, cdr = round(1000 * cdr, 1)),
  rownames = FALSE,
  colnames = c("Year", "CDR per 1000 women"),
  options = list(dom = "T", ordering = FALSE),
  width = "50%"
)
```
<br><br>

Below are age-specific mortality rates for women in the U.S. in 1933 vs. 2019^[Again calculated from the Human Mortality Database].
<details>
<summary>What do you notice?</summary>
Mortality was *up to an order of magnitude higher* in 1933 than 2019 at nearly all ages. Again, WTF is up with CDR comparison?
</details>

```{r us_asmr_paradox_asmr, echo=FALSE, message=FALSE, warning=FALSE}
ggplot2::ggplot(asmr_us) +
  ggplot2::geom_line(aes(x = Age, y = m, color = factor(Year))) +
  ggplot2::annotate(
    "text",
    x = 15,
    y = exp(-5),
    label = "1933",
    color = "#F8766D"
  ) +
  ggplot2::annotate(
    "text",
    x = 30,
    y = exp(-8),
    label = "2019",
    color = "#00BFC4"
  ) +
  ggplot2::scale_y_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  ggplot2::xlab("1-year age group") +
  ggplot2::ylab("Period\nage-specific\nmortality rate\n(log-scale)") +
  ggplot2::theme_minimal() +
  ggplot2::theme(panel.grid = element_blank(),
                 axis.title.y = element_text(angle = 0),
                 legend.position = "none")
```
<br><br>

But remember: $CDR = \sum_{x=0}^{\infty} {}_{n}M_{x}\cdot {}_{n}C_{x}$. Below are the the age structures for women in the U.S. in 1933 vs. 2019^[Again calculated from the Human Mortality Database].
<details>
<summary>What do you notice?</summary>
* Population was older in 2019
* Mortality is higher at older ages
* This could explain things
</details>

```{r us_asmr_paradox_c, echo=FALSE, message=FALSE, warning=FALSE}
ggplot2::ggplot(asmr_us) +
  ggplot2::geom_line(aes(x = Age, y = p, color = factor(Year))) +
  ggplot2::annotate(
    "text",
    x = 30,
    y = 0.0175,
    label = "1933",
    color = "#F8766D"
  ) +
  ggplot2::annotate(
    "text",
    x = 60,
    y = 0.0145,
    label = "2019",
    color = "#00BFC4"
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent) +
  ggplot2::xlab("1-year age group") +
  ggplot2::ylab("Proportion in\nage group") +
  ggplot2::theme_minimal() +
  ggplot2::theme(panel.grid = element_blank(),
                 axis.title.y = element_text(angle = 0),
                 legend.position = "none")
```

## Age-standardization

```{r ascdr_setup, echo=FALSE, message=FALSE, warning=FALSE}
pavg <- asmr_us %>%
  dplyr::group_by(Age) %>%
  dplyr::summarize(p_avg = mean(p)) %>%
  dplyr::ungroup()
p2019 <- asmr_us %>%
  dplyr::filter(Year == 2019) %>%
  dplyr::select(Age, p) %>%
  dplyr::rename(p_2019 = p)
p1933 <- asmr_us %>%
  dplyr::filter(Year == 1933) %>%
  dplyr::select(Age, p) %>%
  dplyr::rename(p_1933 = p)
asmr_standards <- asmr_us %>%
  dplyr::left_join(p2019) %>%
  dplyr::left_join(p1933) %>%
  dplyr::left_join(pavg) %>%
  dplyr::mutate(mp_s2019 = m * p_2019,
                mp_s1933 = m * p_1933,
                mp_savg = m * p_avg)
ascdr <- asmr_standards %>%
  dplyr::group_by(Year) %>%
  dplyr::summarize_at(vars(mp_s2019, mp_s1933, mp_savg), sum) %>%
  dplyr::ungroup()
```

Population aging distorts crude death rate comparisons

**Question:** How can we eliminate the effect of age distribution?

**Answer:** Age-standardized rates!

**Example:** Assume that 1933 instead had a 2019 age structure:^[This is what they call a "counterfactual" in causal inference!]

$$CDR^* = \sum_i M_i^{1933} \cdot C_i^{2019}$$

<details>
<summary>Why not use a 1933 standard? Or average $C_i$ between 1933 and 2019?</summary>
* Choice between 1933 and 2019 is arbitrary, but can largely impact the comparison
* Arguably better to average across age distributions
* But be careful that one extreme age distribution isn't skewing the average
</details>
<br>

**Example again:** Below are standardized and un-standardized CDRs for women in the U.S. in 1933 vs. 2019.
<details>
<summary>What do you notice?</summary>

* CDR difference is smaller for older 2019 standard than younger 1933 standard
* The cross-period average standardized CDR is between those values
* Standardized differences much larger than un-standardized difference
</details>

```{r ascdr_comparison_us, echo=FALSE, warning=FALSE, message=FALSE}
ascdr_diffs <- ascdr %>%
  dplyr::mutate_at(
    vars(starts_with("mp")),
    ~(. - dplyr::lead(.))
  ) %>%
  dplyr::filter(!is.na(mp_s2019)) %>%
  dplyr::select(starts_with("mp")) %>%
  dplyr::mutate(field = "Difference")
sketch <- htmltools::withTags(table(
  class = "display",
  thead(
    tr(
      th(rowspan = 3, ""),
      th(colspan = 4, "CDR estimates per 1000 women")
    ),
    tr(
      th(colspan = 3, "Age structure standard"),
      th(rowspan = 2, "Un-standardized")
    ),
    tr(
      th("1933"),
      th("2019"),
      th("Average")
    )
  )
))
ascdr %>%
  dplyr::rename(field = Year) %>%
  dplyr::mutate(field = as.character(field)) %>%
  dplyr::bind_rows(ascdr_diffs) %>%
  dplyr::bind_cols(tibble::tibble(
    m = c(cdr_us$cdr, cdr_us$cdr[1] - cdr_us$cdr[2])
  )) %>%
  dplyr::mutate_at(vars(starts_with("m")), ~(1000 * .)) %>%
  DT::datatable(
    container = sketch,
    rownames = FALSE,
    options = list(dom = "t", ordering = FALSE)
  ) %>%
  DT::formatStyle(columns = 1, fontWeight = "bold") %>%
  DT::formatRound(columns = 2:5, digits = 1)
```
<br>

More generally, for any standard population $s$ and focal population $j$, the **age-standardized crude death rate** (ASCDR):

$$ASCDR^j = \sum_i M_i^{s} \cdot C_i^{j}$$
Even more generally, for any rate $R$ and schedule of age-specific rates $R_i$:

$$ASR^j = \sum_i R_i^{s} \cdot C_i^{j}$$

## Decomposition of differences between rates or proportions

As a demographer or data scientist, you'll get these questions a lot:

* **Demography:** "How much of the difference between these death rates is due to differences in age distributions?"
* **People analytics:** "How much of the difference between these turnover rates is due to different tenure distributions?"
* **Marketing:** "How much of the customer conversion rate between these two lead generating websites is due to a difference in marketing channel?"

One way to answer such questions is with a clever difference decomposition.

**Example:** Crude death rate difference

$$
\Delta = CDR^B - CDR^A = \sum_i C_i^B \cdot M_i^B - \sum_i C_i^A \cdot M_i^A
$$
On pg. 28, PHG derive a useful decomposition of this difference:

$$
\begin{align}
\Delta &= \sum_i \left(C_i^B-C_i^A\right) \cdot \left[\frac{M_i^B-M_i^A}{2}\right] + \sum_i \left(M_i^B-M_i^A\right) \cdot \left[\frac{C_i^B+C_i^A}{2}\right] \\
       &= \begin{pmatrix}
        \textsf{difference in age} \\
        \textsf{composition}
       \end{pmatrix}
       \cdot \begin{pmatrix}
        \textsf{weighted by avg} \\
        \textsf{age-specific mortality}
       \end{pmatrix} \\
      &+ \begin{pmatrix}
        \textsf{difference in} \\
        \text{rate schedules}
       \end{pmatrix}
       \cdot \begin{pmatrix}
        \textsf{weighted by} \\
        \textsf{average age} \\
        \textsf{composition}
       \end{pmatrix}
\end{align}
$$

Or, put even more simply:

$$
\Delta = \begin{pmatrix}
  \textsf{contribution of} \\
    \textsf{age composition} \\
    \textsf{differences to } \Delta
  \end{pmatrix}
  + \begin{pmatrix} \\
    \textsf{contribution of} \\
    \textsf{rate schedule} \\
    \textsf{differences to } \Delta
  \end{pmatrix}
$$

**Example:** Decomposing the difference in U.S. women's mortality between 1933 vs. 2019. Below is a table of period age-specific population proportions $C_i^\textsf{year}$ and mortality rates $M_i^\textsf{year}$, as well as difference contributions $\left(C_i^{1933}-C_i^{2019}\right) \cdot \left(M_i^{1933}-M_i^{2019}\right)/2$ and $\left(M_i^{1933}-M_i^{2019}\right) \cdot \left(C_i^{1933}-C_i^{2019}\right)/2$.

```{r cdr_us_decomposition_setup, echo=FALSE, message=FALSE, warning=FALSE}
contribs <- asmr_us %>%
  dplyr::select(Year, Age, m, p) %>%
  tidyr::pivot_wider(
    id_cols = Age,
    names_from = Year,
    values_from = c(m, p)
  ) %>%
  dplyr::mutate(
    p_contrib = 0.5 * (p_1933 - p_2019) * (m_1933 + m_2019),
    m_contrib = 0.5 * (m_1933 - m_2019) * (p_1933 + p_2019)
  ) %>%
  dplyr::select(
    Age,
    p_1933, p_2019,
    m_1933, m_2019,
    p_contrib, m_contrib
  )
sum_contribs <- contribs %>%
  dplyr::summarize_all(sum) %>%
  dplyr::select(ends_with("contrib"))
sketch <- htmltools::withTags(table(
  class = "display",
  thead(
    tr(
      th(rowspan = 2, "Age"),
      th(colspan = 2, "Proportion in population"),
      th(colspan = 2, "Mortality rate"),
      th(colspan = 2, "Contribution to difference")
    ),
    tr(
      lapply(rep(c("1933", "2019"), 2), th),
      lapply(rep(c("Age structure", "Rate schedule"), 1), th)
    )
  )
))
DT::datatable(
  contribs,
  container = sketch,
  rownames = FALSE,
  options = list(scrollX = TRUE, dom = "lpt", ordering = FALSE)
) %>%
  DT::formatRound(columns = 2:5, digits = 4) %>%
  DT::formatRound(columns = 6:7, digits = 6)
```

<br><br>

To yield the total contributions of age structure and rate schedule, simply sum up the contribution columns across age groups.^[This is a good example of how looking at the code for this book chapter in GitHub can help you with your problem sets.].
<details>
<summary>What do you notice?</summary>
* Age structure contribution is negative because 1933 population is younger
* Rate schedule contribution is positive because 1933 mortality was higher
</details>

```{r cdr_us_decomposition, echo=FALSE, warning=FALSE, message=FALSE}
sketch <- htmltools::withTags(table(
  class = "display",
  thead(
    tr(
      th(colspan = 2, "Total contribution to difference by...")
    ),
    tr(
      th("Age structure"),
      th("Rate schedule")
    )
  )
))
sum_contribs %>%
  dplyr::mutate_all(~(. * 1000)) %>%
  DT::datatable(
    caption = "Figures per 1000 women",
    container = sketch,
    rownames = FALSE,
    options = list(scrollX = TRUE, dom = "t", ordering = FALSE),
    width = "50%"
  ) %>%
  DT::formatRound(columns = 1:2, digits = 1)
```
<br>

 The sum of these two contributions yields back the original difference:

$$
\Delta = CDR^{1933} - CDR^{2019} = `r round(sum(1000 * as.numeric(sum_contribs)), 1)` \textsf{ per 1000 women}
$$

## The Lexis diagram

:::{.rmdimportant}
**KEY CONCEPT**

The **Lexis diagram** visualizes relationships among:

* Age groups
* Cohorts
* Periods
:::

### Using a Lexis diagram to find three types of age-specific rate {.unnumbered}

Before drawing our first Lexis diagram, let's define three types of rate:

* [Period age-specific rates] (we already defined these early in the week)
* Cohort age-specific rates

$$
{}_{n}R_{x}^{c} = \frac{
  \textsf{Number of occurrences in cohort } c \textsf{ between ages } x \textsf{ and } x + n
}{
  \textsf{Number of person-periods lived in cohort } c \textsf{ between ages } x \textsf{ and } x + n
}
$$

* Cohort period rates

$$
R^{c}[0,T] = \frac{
  \textsf{Number of occurrences in cohort } c \textsf{ between times } 0 \textsf{ and } T
}{
  \textsf{Number of person-periods lived in cohort } c \textsf{ between times } 0 \textsf{ and } T
}
$$
<br>

<details>
<summary>PHG say cohort period rates are "rarely encountered", but can you think of a good example use case?</summary>

:::{.rmdtip}
**DEMOGRAPHY & DATA SCIENCE**

## Cohort period tenure-specific termination rates {.unnumbered}

**A question I've actually heard asked:** "What was the termination rate during peak retail season 2020 (October through December) among people hired in the U.S. during the first surge of the COVID-19 pandemic?"
:::

</details>
<br>

<details>
<summary>Why doesn't a cohort period age-specific rate make any sense?</summary>
For a particular period, a cohort can't be any other ages than the ages they were during that period!

**Example:** Alas, as a member of the 1984^[The year the original *Ghostbusters* was released] birth cohort, I can't experience the age of 25 (ah, youth!) during this period of my life (ah, gainful employment!)
</details>
<br>

```{r lexis_rates_setup, echo=FALSE, warning=FALSE, message=FALSE}
rate_lexis <- LexisPlotR::lexis_grid(
  year_start = 2009,
  year_end = 2021,
  age_start = 0,
  age_end = 12
) +
  ggplot2::xlab("Exact year") +
  ggplot2::ylab("Exact age\nin years") +
  ggplot2::theme(axis.text.x = element_text(angle = 45),
                 axis.title.y = element_text(angle = 0))
label_lexis <- function(.gg, subtitle) {
  .gg +
    ggplot2::labs(
      title = "Lexis diagram for years 2009-2021 and ages 0-12",
      subtitle = subtitle
    )
}
draw_2018 <- function(.gg) {
  .gg %>%
    LexisPlotR::lexis_year(
      lg = .,
      year = 2018,
      fill = "red",
      alpha = 0.25
    )
}
draw_6yo <- function(.gg) {
  .gg %>%
    LexisPlotR::lexis_age(
      lg = .,
      age = 6,
      fill = "blue",
      alpha = 0.25
    )
}
draw_2009c <- function(.gg) {
  .gg %>%
    LexisPlotR::lexis_cohort(
      lg = .,
      cohort = 2009,
      fill = "yellow",
      alpha = 0.25
    )
}
rate_lexis_period <- rate_lexis %>%
  draw_2018() %>%
  label_lexis(
    "Highlighted: Person-years and occurrences counted for all ages in 2018"
  )
rate_lexis_age <- rate_lexis %>%
  draw_6yo() %>%
  label_lexis(
    paste(
      "Highlighted: Person-years and occurrences counted for all people aged 6-7",
      "at any time between 2018 and 2021",
      sep = "\n"
    )
  )
rate_lexis_cohort <- rate_lexis %>%
  draw_2009c() %>%
  label_lexis(
    "Highlighted: My kid's 2009 birth cohort"
  )
rate_lexis_period_age <- rate_lexis %>%
  draw_2018() %>%
  draw_6yo() %>%
  label_lexis(
    "Intersection: Person-years and occurrences for period age-specific rate"
  )
rate_lexis_cohort_age <- rate_lexis %>%
  draw_6yo() %>%
  draw_2009c() %>%
  label_lexis(
    "Intersection: Person-years and occurrences for cohort age-specific rate"
  )
rate_lexis_period_cohort <- rate_lexis %>%
  draw_2018() %>%
  draw_2009c() %>%
  label_lexis(
    "Intersection: Person-years and occurrences for cohort period rate"
  )
rate_lexis_all_together <- rate_lexis %>%
  draw_2018() %>%
  draw_6yo() %>%
  draw_2009c() %>%
  label_lexis("Putting it all together with labels like PHG Figure 2.1") +
  ggplot2::annotate(
    "label", x = as.Date("2018-06-15"), y = 6.5, label = expression("A - P")
  ) +
  ggplot2::annotate(
    "label", x = as.Date("2016-01-01"), y = 6.5, label = expression("A - C")
  ) +
  ggplot2::annotate(
    "label", x = as.Date("2018-06-15"), y = 9, label = expression("C - P")
  )
highlight_plots <- list(rate_lexis_period,
                        rate_lexis_age,
                        rate_lexis_cohort)
intersection_plots <- list(rate_lexis_period_age,
                           rate_lexis_cohort_age,
                           rate_lexis_period_cohort,
                           rate_lexis_all_together)
svg_it <- function(gg)  svglite::xmlSVG(show(gg), standalone = TRUE)
highlight_svg <- highlight_plots %>% purrr::map(svg_it)
intersection_svg <- intersection_plots %>% purrr::map(svg_it)
```
<br><br>

Below I show you how to find periods, cohorts, and age groups in a Lexis diagram

```{r lexis_rates_highlight_carousel, echo=FALSE, warning=FALSE, message=FALSE}
slickR::slickR(highlight_svg, height = 500, width = "95%")
```

Below I show you how to us a Lexis diagram to find the person-years and occurrences to calculate a particular type of rate

```{r lexis_rates_intersection_carousel, echo=FALSE, warning=FALSE, message=FALSE}
slickR::slickR(intersection_svg, height = 500, width = "95%")
```

## Age-specific probabilities

As for rates, probabilities can be restricted to a certain age range^[In fact, we saw in week 1 why most meaningful probabilities in demography are time-bound]

General definition of an age-specific probability:

$$
{}_{n}q_x^{c} = \frac{
  \textsf{Number of occurrences among cohort } c \text{ between ages } x \textsf{ and } x+n
}{
  \textsf{Number of persons in cohort } c \text{ who survived to age } x
}
$$

<details>
<summary>Why is a cohort specified in this definition?</summary>
Because probabilities by definition are cohort measures (see [Probabilities of occurrence of events] in [Concepts and measures])
</details>
<br>

<details>
<summary>For events that can re-occur, what's the problem with this definition?</summary>
Any re-occurring event can happen more than once within a given time range, which makes it possible for $q > 1$, which violates the definition of a probability^[Quantum weirdness aside].

This problem exists for any probability, not just age-specific ones.
</details>
<br>

<details>
<summary>How could you fix that problem?</summary>
Specify that the occurrences counted in the numerator are the *next* occurrence and *only* the next occurrence that any person in the denominator experiences
</details>
<br>

### Using the Lexis diagram to visualize age-specific probabilities {.unnumbered}

```{r lexis_p_setup, echo=FALSE, warning=FALSE, message=FALSE}
lifelines_2009c <- tibble::tribble(
  ~entry,       ~exit,
  "2009-02-14", NA_character_,
  "2009-03-15", "2009-07-04",
  "2009-06-30", NA_character_,
  "2009-08-19", "2010-03-15",
  "2009-09-23", "2010-12-14",
  "2009-10-30", NA_character_,
) %>%
  dplyr::mutate_all(as.Date) %>%
  dplyr::mutate(
    age_at_death = lubridate::interval(entry, exit) / lubridate::years(1)
  )
p_lexis <- LexisPlotR::lexis_grid(
  year_start = 2009,
  year_end = 2014,
  age_start = 0,
  age_end = 5
) +
  ggplot2::xlab("Exact year") +
  ggplot2::ylab("Exact age\nin years") +
  ggplot2::theme(axis.text.x = element_text(angle = 45),
                 axis.title.y = element_text(angle = 0))
draw_lexis_lines <- function(.gg) {
  .gg %>%
    LexisPlotR::lexis_lifeline(
      lg = .,
      birth = lifelines_2009c$entry,
      exit = lifelines_2009c$exit,
      lwd = 1,
      lineends = TRUE
    )
}
q_01 <- lifelines_2009c %>%
  dplyr::ungroup() %>%
  dplyr::summarize(
    d = sum(!is.na(exit[age_at_death < 1])),
    n = dplyr::n()
  ) %>%
  dplyr::mutate(q = scales::percent(d/n, accuracy = 1L))
```

It's pretty easy,actually:

* **Step 1:** Number of lifelines in cohort is denominator of ${}_{n}q_x^c$
* **Step 2:** Number of lifelines that end before age $x+n$ is numerator

<details>
<summary>**Toy example:** Calculate ${}_{1}q_0^{2009c}$ from the Lexis diagram below</summary>

$$
{}_{1}q_0^{2009c}
  = \frac{`r q_01[["d"]]`}{`r q_01["n"]`}
  \approx \textsf{`r q_01[["q"]]`}
$$
</details>

```{r lexis_p, echo=FALSE, message=FALSE, warning=FALSE}
p_lexis %>%
  draw_2009c() %>%
  draw_lexis_lines()
```

<details>
<summary>Why did we exclude one death from the numerator of ${}_{1}q_0^{2009c}$?</summary>
Because it occurred after age $x + n = 0 + 1 = 1$
</details>
<br>

### How the Lexis diagram reveals a key requirement for calculating ${}_{1}q_x^{c}$ from annual birth and death counts {.unnumbered}

Here's some notation to separate deaths between current and previous years:

* ${}_{S}D_x(y)$: Number of deaths at age $x$ in year $y$ among those who reached age $x$ in the **current** year
* ${}_{P}D_x(y)$: Number of deaths at age $x$ in year $y$ among those who reached age $x$ in the **previous** year
* $B_x(y)$: Number of $x$^th^ birthdays in year $y$

:::{.rmdimportant}
**KEY CONCEPT**

To calculate ${}_{1}q_x^{c}$ from death counts for a cohort defined by the year $c$ when it reached age $x$, you need both ${}_{S}D_x(c)$ and ${}_{P}D_x(c+1)$:

$$
{}_{1}q_x^{c} = \frac{{}_{S}D_x(c) + {}_{P}D_x(c + 1)}{B_x(c)} 
$$
:::

**Application:** Probability of death before reaching age 1 for birth cohort $c$:

$$
{}_{1}q_0^{c} = \frac{{}_{S}D_0(c) + {}_{P}D_0(c + 1)}{B_0(c)} 
$$ 

**Example:** Probability of death before reaching age 1 for 2019 birth cohort

$$
{}_{1}q_0^{2009c}
  = \frac{{}_{S}D_0(2009) + {}_{P}D_0(2010)}{B_0(2009)}
  = \frac{1+1}{6}
  = \frac{2}{6}
  \approx \textsf{33%}
$$ 

The Lexis diagram below shows the areas between the 2019 birth cohort's defining lines where relevant births and deaths are counted.

```{r lexis_death_req_setup, echo=FALSE, message=FALSE, warning=FALSE, fig.asp=1}
tibble::tibble(x = seq(from = 2009, to = 2011, by = 0.001),
               y1 = seq(from = 0, to = 2, by = 0.001),
               y2 = y1 - 1) %>%
  ggplot2::ggplot(aes(x = x)) +
  ggplot2::coord_cartesian(xlim = c(2009, 2011), ylim = c(0, 2)) +
  ggplot2::geom_line(aes(y = y1)) +
  ggplot2::geom_line(aes(y = y2)) +
  ggplot2::annotate(
    "text", x = 2009.5, y = 0.1, size = 5,
    label = latex2exp::TeX("$B_0(2009) = 6$")
  ) +
  ggplot2::annotate(
    "text", x = 2009.75, y = 0.25, size = 5,
    label = latex2exp::TeX("${}_{S}D_0(2009) = 1$")
  ) +
  ggplot2::annotate(
    "text", x = 2010.3, y = 0.75, size = 5,
    label = latex2exp::TeX("${}_{P}D_0(2010) = 1$")
  ) +
  ggplot2::geom_ribbon(
    aes(ymin = y2, ymax = y1), fill = "yellow", alpha = 0.25
  ) +
  ggplot2::scale_x_continuous(breaks = c(2009, 2010, 2011), expand = c(0, 0)) +
  ggplot2::scale_y_continuous(breaks = c(0, 1, 2), expand = c(0, 0)) +
  ggplot2::xlab("Exact year") +
  ggplot2::ylab("Exact age\nin years") +
  ggplot2::theme_minimal() +
  ggplot2::theme(panel.grid.minor = element_blank(),
                 axis.title.y = element_text(angle = 0))
```


## Probabilities of death based on mortality experience of a single calendar year

This is a weird section of PHG because:

* The first half is about a measure that isn't all that useful or common
* The second half is about a commonly used measure that is bogus
<br><br>

Let's focus on that second point:

### The so-called infant mortality rate (IMR) is bogus {.unnumbered}

$IMR = \frac{D_0(y)}{B_0(y)}=\frac{{}_{S}D_0(y) + {}_{P}D_0(y)}{B_0(y)}$

<details>
<summary>IMR fails as a rate because its denominator is a poor approximation of infant person-years. Why?</summary>
Counts all births in a year without weighting by person-years lived. A baby who lived one day in year $y$ counted the same as a baby born January 1
</details>
<br>

<details>
<summary>IMR fails as a probability because its numerator includes deaths irrelevant to birth cohort $y$. Eh... why?</summary>
Deaths ${}_{P}D_0(y)$ of infants born last year aren't attributable to births $B_0(y)$ this year
</details>
<br>

$IMR(y) = {}_{1}q_0^y$ if and only if ${}_{P}D_0(y) = {}_{P}D_0(y + 1)$.
<details>
<summary>Most often, ${}_{P}D_0(y) \neq {}_{P}D_0(y + 1)$ Why not?</summary>
Reasons include but are not limited to:

* Growing population, thus more births, thus more infant deaths year to year
* Shrinking population, thus fewer births, thus fewer infant deaths year to year
* Socio-environmental shocks
* Public health improving or declining over time
* Chance year-to-year variation in death counts
</details>
