# Life tables and single decrement processes

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(kableExtra)
library(fuzzyjoin)
library(ggplot2)
options(knitr.kable.NA = "")
```

:::{.rmdimportant}
**KEY CONCEPT**

A **life table** is one of the most important demographic tools for summarizing the mortality experience of a cohort.
:::

### Cohort life tables with lifeline data {.unnumbered}

Since a life table summarizes the experience of a cohort, let's start by building a life table from the lifelines of a small hypothetical cohort based actual people born January 1, 1800^[This section uses the example from PHG pgs. 39-41].

```{r cohort_lt_setup, echo=FALSE, message=FALSE, warning=FALSE}
age_groups <- tibble::tibble(
  x = c(0, 1, 5, 10, 20, 30, 40, 50, 60, 70, 80),
  n = c(1, 4, 5, 10, 10, 10, 10, 10, 10, 10, NA),
  xpn = x + n
) %>%
  dplyr::mutate(group_id = dplyr::row_number())
cohort_data <- tibble::tibble(
  exit_age = c(
    71.55,
    1.22,
    62.91,
    59.60,
    0.07,
    22.12,
    71.14,
    16.41,
    64.05,
    76.79
  )
) %>%
  dplyr::mutate(id = dplyr::row_number())
cohort_lt <- age_groups %>%
  fuzzyjoin::fuzzy_left_join(
    cohort_data,
    by = c("x" = "exit_age", "xpn" = "exit_age"), 
    match_fun = list(`<=`, `>`)
  ) %>%
  dplyr::mutate(indiv_A = dplyr::case_when(
    (exit_age < x) ~ 0,
    (x <= exit_age) & (exit_age < xpn) ~ exit_age - x,
    exit_age >= xpn ~ n
  )) %>%
  dplyr::group_by(x, n,xpn) %>%
  dplyr::summarize(
    d = sum(!is.na(exit_age)),
    A = sum(indiv_A),
    a = mean(indiv_A)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    l = nrow(cohort_data) - cumsum(dplyr::lag(d, default = 0)),
    q = d / l,
    p = 1 - q,
    L = (l - d) * n + ifelse(is.na(A), 0, A),
    m = d / L,
    T = rev(L) %>% dplyr::coalesce(0) %>% cumsum() %>% rev(),
    e0 = T / l
  )
cohort_lt_xL <- cohort_lt %>% dplyr::select(x, L)
```

Below, a lifelines plot like we drew in [week 1](#person-periods), except time is measured from age $x = 0$, and everybody in this real birth cohort was born January 1, 1800^[From PHG Figure 3.1]:

```{r cohort_lifelines_plot, echo=FALSE, message=FALSE, warning=FALSE}
cohort_data %>%
  ggplot2::ggplot() +
  ggplot2::aes(x = 0, xend = exit_age, y = factor(rev(id)), yend = ..y..) +
  ggplot2::geom_segment(lwd = 1) +
  ggplot2::geom_text(aes(x = exit_age, label = exit_age), hjust = -1) +
  ggplot2::scale_x_continuous(
    breaks = c(0, 1, 5, seq(10, 80, by = 10)), expand = c(0, 0)
  ) +
  ggplot2::coord_cartesian(xlim = c(0, 80), clip = "off") +
  ggplot2::xlab("Exact age (years)") +
  ggplot2::ylab(NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_line(),
    axis.line.x = element_line(),
    axis.text.y = element_blank(),
    plot.margin = unit(c(1, 3, 1, 1), "lines")
  )
```
<br><br>

<details>
<summary>
Number $l_0$ of live births (i.e., number surviving to exact age $x = 0$)?</summary>
$l_{0} = `r cohort_lt[["l"]][cohort_lt[["x"]] == 0]`$
</details>

<details>
<summary>Number of deaths ${}_{n}d_{x}$ between ages $x = 0$ and $x+n = 1$?</summary>
${}_{1}d_{0} = `r cohort_lt[["d"]][cohort_lt[["x"]] == 1]`$
</details>

<details>
<summary>
Number $l_x$ still alive at exact age $x = 1$?</summary>
$l_{1} = l_0 - {}_{1}d_{0} = `r cohort_lt[["l"]][cohort_lt[["x"]] == 1]`$

So in general:

$$l_{x+n} = l_x - {}_{n}d_{x}$$
</details>

<details>
<summary>
Probability of death ${}_{n}q_{x}$ between ages $x = 0$ and $x+n = 1$?</summary>
$$
{}_{1}q_{0} = \frac{{}_{1}d_{0}}{l_0}
             = `r cohort_lt[["q"]][cohort_lt[["x"]] == 0]`
$$

So in general:

$${}_{n}q_{x} = \frac{{}_{n}d_{x}}{l_x}$$
</details>

<details>
<summary>
Probability of survival ${}_{n}p_{x}$ between ages $x = 0$ and $x+n = 1$?</summary>
$$
{}_{1}p_{0} = l_1 / l_0
              = 1 - {}_{1}q_{0}
              = `r cohort_lt[["p"]][cohort_lt[["x"]] == 0]`
$$

For general $x$:

$${}_{n}p_{x} = l_{x+n} / l_x = 1 - {}_{n}q_{x}$$
</details>

<details>
<summary>Number of person-years ${}_{n}A_x$ lived by those who died between ages $x$ and $x+n$?</summary>
For general $x$:

$$
{}_{n}A_x =
  \sum_{i \in \textsf{dead}_x} \left(\textsf{(Exact age at death)}_i - x\right)
$$

But for these $x = 0$ (i.e., infants):

$$\begin{align}
{}_{1}A_0
  &= \sum_{i \in \textsf{dead}_0} \left(\textsf{(Exact age at death)}_i - 0\right) \\
  &= \sum_{i \in \textsf{dead}_0} \textsf{(Exact age at death)}_i \\
  &= `r cohort_lt[["A"]][cohort_lt[["x"]] == 0]`
\end{align}$$
</details>

<details>
<summary>What is the average number of person-years lived ${}_{n}a_x$ by those who died between ages $x$ and $x+n$?<br>
**Hint:** It's the average of a similar figure we've seen already, thus a ratio of that sum to a count we've seen already.</summary>
For general age group starting at exact age $x$:

$${}_{n}a_x = \frac{{}_{n}A_x}{{}_{n}d_x}$$

But for these $x = 0$ (i.e., infants):

$$
{}_{1}a_0
  = \frac{{}_{1}A_0}{{}_{1}d_0}
  = \frac{
    `r cohort_lt[["A"]][cohort_lt[["x"]] == 0]`
  }{
    `r cohort_lt[["d"]][cohort_lt[["x"]] == 0]`
  }
  = `r cohort_lt[["a"]][cohort_lt[["x"]] == 0]`
$$
</details>

<details>
<summary>
Person-years lived ${}_{n}L_{x}$ between ages $x = 0$ and $x+n = 1$?</summary>
$$\begin{align}
  {}_{n}L_{x}
    &= \begin{pmatrix}
      \textsf{Person-years lived} \\
      \textsf{between ages } x \textsf{ and } x+n \\
      \textsf{among survivors to age } x+n
    \end{pmatrix}
    \times \begin{pmatrix}
      \textsf{Width of } [x,x+n)\\
      \textsf{age interval}
    \end{pmatrix}
    + \begin{pmatrix}
      \textsf{Person-years lived} \\
      \textsf{between ages } x \textsf{ and } x+n \\
      \textsf{among those who died} \\
      \textsf{before age } x+n
    \end{pmatrix} \\
    &=  \begin{pmatrix}
      \textsf{Survivor count} \\
      \textsf{weighted by} \\
      \textsf{interval width} \\
    \end{pmatrix}
    + \begin{pmatrix}
      \textsf{Fractional person-years} \\
      \textsf{lived by the now deceased}
    \end{pmatrix} \\
    &= \left(l_{x+n} \times n\right) + {_n}_{}A_x
\end{align}$$

This should look familiar to our expression of person-years from [week 1](#person-periods): $PY[0,1] = \sum_1^P N_i \times \Delta_i$

So for these 1800 birth cohort infants:

$$\begin{align}
{}_{1}L_0
  &= l_1 \times 1
  + \sum_{i \in \textsf{dead}_0} \left(\textsf{(Exact age at death)}_i - 0\right) \\
  &= l_1  + \sum_{i \in \textsf{dead}_0} \textsf{(Exact age at death)}_i \\
  &= `r cohort_lt[["l"]][cohort_lt[["x"]] == 1]` + `r cohort_lt[["A"]][cohort_lt[["x"]] == 0]` \\
  &= `r cohort_lt[["l"]][cohort_lt[["x"]] == 1] + cohort_lt[["A"]][cohort_lt[["x"]] == 0]`
\end{align}$$

Recall that ${}_{n}a_x = \frac{{}_{n}A_x}{{}_{n}d_x}$

Thus ${}_{n}A_x = {}_{n}a_x \times {}_{n}d_x$

Thus $L_x = l_{x+n} \times n + {_n}_{}a_x \times {}_{n}d_x$

This becomes useful when doing **period life tables** because:

* You can't calculate ${}_{n}A_x$ directly
* So you can't calculate $L_x$ from ${}_{n}A_x$ directly
* So you borrow values from a model or another population
* But you want ${}_{n}A_x$ estimates that aren't contaminated by another population's size
<br><br>
</details>

<details>
<summary>Person-years $T_x$ lived above exact age $x = 0$?</summary>
$$\begin{align}
  T_0
    &= \textsf{Total person years lived from birth to death} \\
    &= \sum_i \textsf{(Exact age at death)}_i \\
    &= \sum_{a=0}^\infty {}_{n}L_a
\end{align}$$

:::{.rmdtip}
**DEMOGRAPHY & DATA SCIENCE**
<details>
<summary>Okay that was too easy, what about $T_{20}$? Let's do it with R code.</summary>
More generally:

$$T_x = \sum_{a=x}^\infty {}_{n}L_a$$

Easy to think about, but funky to do in practice.

Let's walk through how you'd do it with R code.

Say you have a cohort life table in a `data.frame` with two columns:

* `x`: Exact age $x$ at the beginning of each age interval
* `L`: Person-years ${}_{n}L_x$ lived in the age interval

It looks like this:

```{r cohort_lt_xL_display, echo=FALSE, message=FALSE, warning=FALSE}
cohort_lt_xL %>%
  knitr::kable() %>%
  kableExtra::kable_paper("hover", full_width = FALSE, position = "left")
```

To create a column `T` storing $T_x$ values, do this (assuming you have `dplyr` installed and loaded):

```{r cohort_lt_xL_create_T, message=FALSE, warning=FALSE}
saving_you_some_time <- cohort_lt_xL %>%
  dplyr::mutate(
    T = rev(L) %>% # Reverse person-years column so sum back to x = 0
      dplyr::coalesce(0) %>% # Replace the missing value at x = 80 so you can...
      cumsum() %>% # ... take the cumulative sum of person years
      rev() # Finally, reverse the cumulative sum back in age order
  )
saving_you_some_time %>%
  knitr::kable() %>%
  kableExtra::kable_paper("hover", full_width = FALSE, position = "left")
```

You're welcome.
</details>

:::
</details>

<details>
<summary>What is the life expectancy at birth, i.e., life expectancy $e_x^0$ at age $x = 0$?<br>
**Hint:** It's the per-survivor average of a value we've already calculated, so it's a ratio of that value and another one we've already calculated.</summary>
$$\begin{align}
  e_0^0
    &= \frac{
      \textsf{Person-years lived after age } x=0 \textsf{ (i.e., total person-years)}
    }{
      \textsf{People alive after age } x=0 \textsf{ (i.e., live births)}
    } \\
    &= T_0/l_0
    = `r cohort_lt[["T"]][cohort_lt[["x"]] == 0]`
    / `r cohort_lt[["l"]][cohort_lt[["x"]] == 0]`
    = `r cohort_lt[["e0"]][cohort_lt[["x"]] == 0]`
\end{align}$$

So for general exact age $x$:

$$e_x^0 = T_x/l_x$$

</details>

<details>
<summary>What is the mortality rate ${}_{n}m_x$ in the cohort between ages $x$ and $x+n$</summary>
* This is just a cohort age-specific rate
* Using lifetable notation, it is:

$$
{}_{1}m_0
  = {}_{1}d_0 / {}_{1}L_0
  = `r cohort_lt[["d"]][cohort_lt[["x"]] == 0]`
  / `r cohort_lt[["L"]][cohort_lt[["x"]] == 0]`
  = `r cohort_lt[["m"]][cohort_lt[["x"]] == 0]`
$$
</details>
<br><br>

<center>`r emo::ji("exploding head")``r emo::ji("exploding head")``r emo::ji("exploding head")`</center>

**QUESTION:** Why did we just learn all of those values?

**ANSWER:** Three reasons:

1. I'm evil
2. Their relationships and values help create **period life tables**
3. Their values (and operations on them) are interesting in and of themselves

* We'll see examples of those last two points in later sections.
* You've already seen ample evidence of the first point.

Now let's put all these values together into our first cohort life table:

```{r cohort_lt_display, echo=FALSE, message=FALSE, warning=FALSE}
cohort_lt_display <- cohort_lt %>%
  dplyr::select(x, n, xpn, l, d, q, p, L, T, e0, m, a) %>%
  setNames(
    nm = c(
      "$x$",
      "$n$",
      "$x+n$",
      "$l_x$",
      "${}_{n}d_x$",
      "${}_{n}q_x$",
      "${}_{n}p_x$",
      "$L_x$",
      "$T_x$",
      "$e_x^0$",
      "${}_{n}m_x$",
      "${}_{n}a_x$"
    )
  )
cohort_lt_display %>%
  knitr::kable() %>%
  kableExtra::kable_paper("hover", full_width = FALSE, position = "left") %>%
  kableExtra::column_spec(2:ncol(cohort_lt_display), border_left = TRUE)
```


## Period life tables

<details>
<summary>Which of the values we've covered so far would be hard to come by in classic demographic data. Why?</summary>
* Total or average person-years lived in age intervals (don't have exact person years)
* Probabilities of death or survival (don't have cohort data or separated count data)
* Person-years lived in age interval (but can estimate using [week 2](#estimating-person-periods) methods)
</details>
<br><br>

## Strategies for ${}_{n}m_x$ $\rightarrow$ ${}_{n}q_x$ conversion

<details>
<summary>Why is converting rates to probabilities important for creating period life tables?</summary>
We often have access to period rates, but not probabilities (cohort measures)
</details>
<br>

### Direction observation (duh)

### Graduation of the age-specific mortality rate (${}_{n}m_x$) function

### Borrowing average person-years lived (${}_{n}a_x$) values from another population

### Using rules of thumb

## The very young ages

## The open-ended age interval

## Review of the steps for period life table construction

## Interpreting the life table

## The life table conceived as a stationary population

## Mortality as a continuous process

## Life table construction revisited

## Decomposing a difference in life expectancies

## Adaptation of the life table for studying other single decrement processes

:::{.rmdtip}
**DEMOGRAPHY & DATA SCIENCE**

### Beware of the all-cause attrition hegemon
:::
